Data Formatting

### 2021-02-19 scripts/data_format.Rmd
clean up code

community summary - includes each site_month as rows and variables cc, size, prevalenceetc... as columns
```{r}
comm_summ <- data %>% dplyr::select(WetAltID, Month.1, cc, Month, AB2:AB42, Prevalence) %>% 
  dplyr::select(-ABAmb, -ABSal) %>%  mutate(., Size = rowSums(.[5:25])) %>% #get totals for community size
  dplyr::select(-AB2:-AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>%
  distinct()
comm_summ$siteID <- gsub("Month", "", comm_summ$siteID)
```

@@

environmental variables matrix
```{r}
#select for environmental variables
env <- data %>% dplyr::select(WetAltID, Month.1, MeanAirT, MeanWaterTempPredC, DryingScore, CanopyCover, Area, Perimeter) %>% distinct() %>% arrange(.,WetAltID) 
env$Month <- gsub("Month", "", env$Month.1)
#make siteID rowname
env %<>% mutate(., siteID = paste(env$WetAltID, env$Month, sep = "_")) %>% distinct() %>% arrange(.,WetAltID) %>% dplyr::select(siteID, MeanAirT:Perimeter)

#need to remove duplicated rows and determine what to do with NA's
env_mat <- env %>% column_to_rownames(., var = "siteID")
```
environmental variables are `r colnames(env_mat)`

@@

richness values for each site
```{r}
#Get species richness from pres/abs matrix
richness <- tibble()
for (i in 1:length(rownames(presence_mat))){
  richness[i,1] <- rownames(presence_mat)[i]
  richness[i,2] <- sum(presence_mat[i,])
}
richness %<>% rename(siteID = 1, richness = 2)
```

@@

  #calculate disease-free equilibria for use in next-gen matrix
  data %<>% add_column(dfe1 = (data$birth1/data$mort1),dfe2 = (data$birth2/data$mort2)) 
  #make matrix
  mat <- matrix(1, nrow = 3, ncol = 3)

  m <- nrow(data)
  #calculate eigenvalue using next-gen matrix. max eigenvalue is R0
  for(n in 1:m){
    mat[1] = (data$trans1[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[2] = (data$trans1[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[3] = (data$shed1[n]/data$degr[n])
    mat[4] = (data$trans2[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[5] = (data$trans2[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[6] = (data$shed2[n]/data$degr[n])
    mat[7] = (data$trans3[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[8] = (data$trans3[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[9] = 0
    eigen <- max(eigen(mat)$values)
    data$eigen[n] <- eigen
  }

@@

cc x evenness data - primarily used by fig3
```{r}
richness_cc <- full_join(comm_summ, richness_mat, by = "siteID")

@@

#remove the first entry for each wetland to remove the carryover from the last wetland
lag_evenness %<>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)

data for bar plots - primarily used by fig4
```{r}

#Get species richness from matrix

community_mat <- data %>% dplyr::select(WetAltID, Month.1, AB2:AB8, AB9, AB20:AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>% distinct() #select data and make siteID column

community_mat <- community_mat[order(community_mat$WetAltID, community_mat$Month.1),] #order by wetland and month

community_mat %<>% arrange(.,WetAltID) %>% dplyr::select(-WetAltID, -Month.1) %>% dplyr::select(siteID, AB2:AB42)

abundances <- community_mat #make abundances df

community_mat %<>% column_to_rownames(., var = "siteID") #make sitexspecies matrix

presence_mat <- ifelse(community_mat>0, 1, 0) #make pres-abs sitexspecies matrix

siteID <- rownames(presence_mat)

richness_mat <- tibble()
for (i in 1:length(siteID)){
  richness_mat[i,1] <- siteID[i]
  richness_mat[i,2] <- sum(presence_mat[i,])
}
richness_mat %<>% rename(siteID = 1, richness = 2) #make df with siteIDs and corresponding species richness

#Get species richness for each site and community competence values

tmp <- data %>% dplyr::select(WetAltID, Month.1, cc, Prevalence,AB2:AB8, AB9, AB20:AB42) %>% 
    distinct() %>% 
    mutate(., siteID = paste(WetAltID, Month.1, sep = "_"))
tmp %<>% mutate(.,size = rowSums(.[5:25])) %>% select(-c(AB2:AB42))
tmp <- tmp[order(tmp$WetAltID, tmp$Month.1),]
tmp %<>% dplyr::select(-Month.1)

richness_cc <- full_join(richness_mat, tmp, by = "siteID") #make df with siteIDs, richness, and cc

h <- vegan::diversity(community_mat, index = "shannon") #calculate shannons index for each site-month
h <- melt(as.matrix(h)) #melt into df
h %<>% rename(siteID = Var1, h = value) %>% dplyr::select(-Var2) #get df with shannons index for each site-month

evenness <- richness_cc %>% mutate(lnS = log(richness)) %>% rowwise() 
evenness %<>% full_join(h, evenness, by = "siteID")
evenness %<>% mutate(J = h/lnS) %>% rowwise() #calculate Pielou's J

#test for prevalence-richness relationship

evenness %>% ggplot(.,aes(x=richness, y=Prevalence))+
  geom_smooth(method="lm")
cor.test(evenness$richness,evenness$Prevalence,method="spearman")

#create new df for ordered sites to test lagged richness and evenness on prevalence
lag_evenness <- evenness[-c(4,14,47,90),] #remove site-months that are out of sequence
lag_evenness$lag_richness <- c(0)
lag_evenness$lag_J <- c(0)
lag_evenness$lag_cc <- c(0)
lag_evenness$lag_size <- c(0)
for(n in 2:91){ #lag values by one month
  lag_evenness$lag_richness[n] <- lag_evenness$richness[n-1]
  lag_evenness$lag_J[n] <- lag_evenness$J[n-1]
  lag_evenness$lag_cc[n] <- lag_evenness$cc[n-1]
  lag_evenness$lag_size[n] <- lag_evenness$size[n-1]
}
#remove the first entry for each wetland to remove the carryover from the last wetland
lag_evenness %<>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)
#lag_evenness %<>% select(siteID, lag_richness:lag_size)

lag_evenness %>% ggplot(.,aes(x=lag_richness, y=Prevalence))+
  geom_smooth(method="lm")
cor.test(lag_evenness$lag_richness,lag_evenness$Prevalence,method="spearman")

#barplots comparing community composition

evenness %<>% select(-WetAltID)
abundances %<>% left_join(.,evenness, by = "siteID")
lag_evenness %<>% select(siteID, lag_cc, lag_richness, lag_J, lag_size)
abundances %<>% inner_join(.,lag_evenness, by = "siteID")
abundances <- mutate(abundances, ABHigh = (AB9 + AB21 + AB26 + AB42)) 
abundances <- mutate(abundances, ABLow = (AB2 + AB3 + AB4 + AB5 + AB6 + AB8 + AB20 + AB24 + AB27 + AB28 + AB29 + AB31 + AB34 + AB35 + AB38 + AB39 + AB41))
abundances <- mutate(abundances, total = ABHigh + ABLow)
#final df for plots. includes data for abundance, prevalence, cc, diversity measure, lagged values

@@

correlations - primarily used by fig5
```{r}
prev_cc <- data %>% dplyr::select(WetAltID, Month.1, cc, Prevalence, Month, AB2:AB42, MeanWaterTempPredC)
prev_cc %<>% dplyr::select(-ABAmb, -ABSal) %>%  mutate(., total = rowSums(.[6:26])) #get totals for community size
prev_cc %<>% dplyr::select(-AB2:-AB42)
prev_cc %<>% distinct()

@@

order$lag_cc <- order$cc
order$lag_size <- order$total
order$lag_temp <- order$MeanWaterTempPredC

#create new column that includes previous month's value for cc
for(n in 2:nrow(order)){
  order$lag_cc[n] <- order$cc[n-1]
  order$lag_size[n] <- order$total[n-1]
  order$lag_temp[n] <- order$MeanWaterTempPredC[n-1]
}

#remove the first entry for each wetland to remove the carryover from the last wetland
clean <- order %>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)

>>>>

community summary - includes each site_month as rows and variables cc, size, prevalenceetc... as columns
```{r}
comm_summ <- data %>% dplyr::select(WetAltID, Month.1, Month, cc, Month, AB2:AB42, Prevalence, MeanWaterTempPredC) %>% 
  dplyr::select(-ABAmb, -ABSal) %>%  mutate(., size = rowSums(.[5:25])) %>% #get totals for community size
  dplyr::select(-AB2:-AB42) %>%
  mutate(., siteID = paste(WetAltID, Month.1, sep = "_")) %>%
  distinct()
comm_summ$siteID <- gsub("Month", "", comm_summ$siteID)
```

@@

environmental variables matrix - needs to be fixed!!
```{r}
#select for environmental variables
env <- data %>% dplyr::select(WetAltID, Month.1, MeanAirT, MeanWaterTempPredC, DryingScore, CanopyCover, Area, Perimeter) %>% distinct() %>% arrange(.,WetAltID) 
env$Month <- gsub("Month", "", env$Month.1)
#make siteID rowname
env %<>% mutate(., siteID = paste(env$WetAltID, env$Month, sep = "_")) %>% distinct() %>% arrange(.,WetAltID) %>% dplyr::select(siteID, MeanAirT:Perimeter)

#need to remove duplicated rows and determine what to do with NA's
#env_mat <- env %>% column_to_rownames(., var = "siteID")
```
environmental variables are `r colnames(env_mat)`

@@

richness values for each site
```{r}
#Get species richness from pres/abs matrix
richness <- tibble(siteID="", richness=1:nrow(presence_mat))
for (i in 1:nrow(presence_mat)){
  richness[i,1] <- rownames(presence_mat)[i]
  richness[i,2] <- sum(presence_mat[i,])
}
```

@@

  #calculate disease-free equilibria for use in next-gen matrix
  data %<>% add_column(dfe1 = (data$birth1/data$mort1),dfe2 = (data$birth2/data$mort2), eigen = NA) 
  #make matrix
  mat <- matrix(1, nrow = 3, ncol = 3)

  m <- nrow(data)

  #calculate eigenvalue using next-gen matrix. max eigenvalue is R0
  for(n in 1:m){
    mat[1] = (data$trans1[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[2] = (data$trans1[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[3] = (data$shed1[n]/data$degr[n])
    mat[4] = (data$trans2[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[5] = (data$trans2[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[6] = (data$shed2[n]/data$degr[n])
    mat[7] = (data$trans3[n]*data$dfe1[n])/(data$reco1[n]+data$mort1[n])
    mat[8] = (data$trans3[n]*data$dfe2[n])/(data$reco2[n]+data$mort2[n])
    mat[9] = 0
    eigen <- max(eigen(mat)$values)
    data$eigen[n] <- eigen
  }

@@

cc x evenness data - primarily used by fig3
```{r}
richness_cc <- full_join(comm_summ, richness, by = "siteID")

@@

#remove the first entry for each wetland to remove the carryover from the last wetland
lag_evenness %<>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)
```

data for bar plots - primarily used by fig4
```{r}
#barplots comparing community composition
abundances <- community_mat %>% left_join(.,evenness, by = "siteID")
lag_evenness %<>% select(siteID, lag_cc, lag_richness, lag_J, lag_size)
abundances %<>% inner_join(.,lag_evenness, by = "siteID")
abundances <- mutate(abundances, ABHigh = (AB9 + AB21 + AB26 + AB42)) 
abundances <- mutate(abundances, ABLow = (AB2 + AB3 + AB4 + AB5 + AB6 + AB8 + AB20 + AB24 + AB27 + AB28 + AB29 + AB31 + AB34 + AB35 + AB38 + AB39 + AB41))
abundances <- mutate(abundances, total = ABHigh + ABLow)
#final df for plots. includes data for abundance, prevalence, cc, diversity measure, lagged values

@@

correlations - primarily used by fig5
```{r}
prev_cc <- comm_summ

@@

order %<>% add_column(lag_cc = NA, lag_size = NA, lag_temp = NA)

#create new column that includes previous month's value for cc
for(n in 2:nrow(order)){
  order$lag_cc[n] <- order$cc[n-1]
  order$lag_size[n] <- order$size[n-1]
  order$lag_temp[n] <- order$MeanWaterTempPredC[n-1]
}

#remove the first entry for each wetland to remove the carryover from the last wetland
clean <- order %>% group_by(WetAltID) %>% filter(duplicated(WetAltID) | n()==1)
```
